Parsing Postfix log files

### 2009-02-26 doc/logparser-architecture.tex
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.

Each action is a separate procedure written to deal with a particular
category of input, e.g.\ rejections.  The actions are parser-specific: each
parser author will need to write the required actions from scratch unless
extending an existing parser.  It is anticipated that parsers based on this
architecture will have a high ratio of rules to actions, with the aim of
having simpler rules and clearer distinctions between the inputs parsed by
different rules.

The ability to add special purpose actions to deal with difficulties and
new requirements that are discovered during parser development is one of
the strengths of this architecture.  Instead of writing a single monolithic
function that must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.
Sometimes the new action will require the cooperation of other actions,
e.g.\ to set or check a flag.  There is a possibility of introducing
failure when modifying existing actions in this way, but the modifications
will be smaller and occur less frequently than with a monolithic
architecture, thus failures will be less likely to occur and will be easier
to test for and diagnose.  The architecture can be implemented in an object
oriented style, allowing sub-classes to extend or override actions in
addition to adding new actions; because each action is an independent
procedure, the sub-class need only modify the action it is overriding,
rather than reproducing large chunks of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log lines there were many complications
to be overcome.  Some were the result of deficiencies in Postfix's logging
(some of which were rectified by later versions of Postfix); others were
due to the vagaries of process scheduling, client behaviour, and
administrative actions.  All were successfully accommodated in the Postfix
log parser: adding new actions was enough to overcome several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
existing actions to cooperate and exchange extra data (via the framework),
changing their behaviour as appropriate based on that extra data.

Actions may return a modified input line that will be parsed as if read
from the input stream, allowing for a simplified version of cascaded
parsing~\cite{cascaded-parsing}.  This powerful facility allows several
rules and actions to parse a single input, potentially simplifying both
rules and actions.

@@

Rules categorise inputs, specifying both the regex to match against each
input and the action to invoke when the match is successful.  Parsing new
inputs is generally achieved by creating a new rule that pairs an existing
action with a new regex.  Decoupling the rules from the actions and
framework enables other rule management approaches to be used, e.g.\
instead of manually adding new rules, machine learning techniques could be
used to automatically generate new rules.  If this approach was taken the
choice of machine learning technique would be constrained by the size of
typical data sets (see \sectionref{Results} XXX CHECK THIS REFERENCE WHEN
THE RESULTS SECTION IS FINISHED).  Techniques requiring the full data set
when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser might also dynamically create new rules in response to certain
inputs, e.g.\ diagnostic messages indicating the program which produced the
input being parsed had read a new configuration file.  These avenues of
research and development has not been pursued by the author, but could
easily be undertaken independently.

The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each input might change between parser invocations.  Overlapping
rules are frequently a requirement, allowing a more specific rule to match
some inputs and a more general rule to match the remainder, e.g.\
separating \acronym{SMTP} delivery to specific sites from \acronym{SMTP}
delivery to the rest of the world.  Allowing overlapping rules simplifies
both the general rule and the more specific rule; additionally rules from
different sources can be combined with a minimum of prior cooperation or
modification required.  Overlapping rules should have a priority attribute
to specify their relative ordering; negative priorities may be useful for
catchall rules.

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
computation complexity.  Perl 5.10 regexes are more powerful than
traditional regexes: it is possible to match correctly balanced brackets
nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Perl 5.10 regexes can maintain an arbitrary
state stack and are thus equivalent in computational power to \acronym{PDA}
or \acronym{CFL}, so detecting overlap may require calculating the
intersection of two \acronym{PDA} or \acronym{CFL}s.  The intersection of
two \acronym{CFL}s is not closed, i.e.\ the resulting language cannot
always be parsed by a \acronym{CFL}, so intersection may be intractable in
some cases e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating intersection.  This is certainly not a task to be
performed every time the parser is used: detecting overlap amongst the
Postfix log parser's \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

The framework requires each rule to have \texttt{action} and \texttt{regex}
attributes; each implementation is free to add any additional attributes it
requires.

It is possible to define pathological regexes which fall into two main
categories: regexes that match every input, and regexes that consume
excessive amounts of CPU time during matching.  Defining a regex to match
all inputs is trivial: \verb!/^/! matches the start of every input.
Usually excessive CPU time is consumed when a regex with a lot of
alteration and variable quantifiers fails to match, but successful
matching is generally quite fast (see~\cite{mastering-regular-expressions}
for in-depth discussion).

XXX RULES CAN HAVE CONDITIONS ATTACHED

\label{comparison against context-free grammars}

In context-free grammar terms the parser rules could be described as:

$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

@@

XXX IMPROVE THIS SECTION\@; SHOULD IT BE A SUBSECTION\@?

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.

    \item The first matching rule wins: no further rules are tried against
        that log line, but there is a mechanism for prioritising the rules
        so that more specific rules can be tried first.

    \item Rules are completely self-contained and can be understood in
        isolation, without reference to any other rules.

    \item Rule processing time is a linear function of the number of rules.
        XXX IMPROVE THIS\@.

@@

@@

    \item [Matching rules against inputs is simple:]  The first matching
        rule determines the action that will be invoked: there is no
        backtracking to try alternate rules, no attempt is made to pick a
        \textit{best\/} rule.

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push unused
        input back onto the input stream.  This was not a deliberate design
        decision, rather a consequence of the line oriented nature of
        Postfix log files; more flexible approaches could be pursued.

    \item [Context-free rules:]  Rules can not take into account past or
        future inputs.  In context-free grammar terms the parser rules
        could be described as:
        \newline{}$\text{\textless{}input\textgreater{}} \mapsto
        \text{rule-1} | \text{rule-2} | \text{rule-3} | \dots |
        \text{rule-n}$.

@@

    \item [Transduction:]  The architecture can be thought of as
        implementing transduction: it takes data in one form (log files)
        and transforms it to another form (a database); other formats may
        be more suitable for other implementations.

    \item [Similarity to \acronym{NLP}:] \hfill{} \newline{}
        Unlike traditional parsers such as those used when compiling a
        programming language, this architecture does not require a fixed
        grammar specification that inputs must adhere to.  The architecture
        is capable of dealing with interleaved inputs, out of order inputs,
        and ambiguous inputs where heuristics must be applied --- all have
        arisen and been successfully accommodated in the Postfix log
        parser.

>>>>

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejections.  There may be many input variants
within one input category; in general each action will handle one category
of inputs, with each rule rule recognising one input variant.  It is
anticipated that parsers based on this architecture will have a high ratio
of rules to actions, with the aim of having simple rules and clear
distinctions between the inputs parsed by different rules.  An action may
process different input variants in slightly different ways, but large
changes in processing indicate the need for a new action and a new category
of input; if an action becomes unnecessarily complicated is starts to turn
into a monolithic parser, with too much logic contained in a single
procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
must be modified to support new behaviour, with all the attendant risks of
adversely affecting the existing parser, when a new requirement arises an
independent action can be written to satisfy it.  Sometimes the new action
will require the cooperation of other actions, e.g.\ to set or check a
flag, or save some extra data.  There is a possibility of introducing
failure when modifying existing actions in this way, but the modifications
will be smaller and occur less frequently than with a monolithic
architecture, thus failures will be less likely to occur and will be easier
to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
implemented in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent procedure, the sub-class need only modify the action it is
overriding, rather than reproducing large chunks of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log \empty{fool check}lines there were
many complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were rectified by later versions of
Postfix, e.g.\ identifying bounce
notifications~\sectionref{identifying-bounce-notifications}); others were
due to the vagaries of process scheduling, client behaviour, and
administrative actions.  All were successfully accommodated in the Postfix
log parser: adding new actions was enough to overcome several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
existing actions to cooperate and exchange extra data (via the framework),
changing their behaviour as appropriate based on that extra data.  Every
architecture should aim to make the hard things possible as well as making
the easy things easy (XXX MAYBE QUOTE LARRY\@?); the successful
implementation of the Postfix log parser demonstrates that this
architecture achieves that aim.

Actions may return a modified input line that will be parsed as if read
from the input stream, allowing for a simplified version of cascaded
parsing~\cite{cascaded-parsing}.  This powerful facility allows several
rules and actions to parse a single input, potentially simplifying both
rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.

@@

XXX DO I NEED TO SUBDIVIDE THIS INTO SUBSECTIONS\@?

Rules are responsible for categorising inputs: each rule should recognise
one and only input variant; an input category with multiple input variants
should have multiple rules, one for each variant.  Rules will typically use
a regex when classifying inputs, but other approaches may prove useful for
some applications, e.g.\ comparing fixed strings to the input, or checking
the length of the input; for the remainder of this thesis it will be
assumed that a regex is used.  A rule must specify both the regex to match
against each input and the action to invoke when the match is successful,
but implementations are free to add any other attributes they require;
\sectionref{rule attributes} describes the attributes used in
\parsername{}, and some potential attributes will be discussed later in
this section.

The framework issues a warning for each unparsed input, so it is evident
when the ruleset needs to be augmented.  Parsing new inputs is achieved in
one of three ways:

\begin{enumerate}

    \item Modify an existing rule's regex, because the new input is part of
        an existing variant.

    \item Write a new rule that pairs an existing action with a new regex;
        this adds a new variant to an existing category.

    \item Create a new category of inputs and a new action to process
        inputs from the new category; write a new rule pairing the new
        action with a new regex.

\end{enumerate}

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually adding new
rules, machine learning techniques could be used to automatically generate
new rules.  If this approach was taken the choice of machine learning
technique would be constrained by the size of typical data sets (see
\sectionref{parser efficiency}).  Techniques requiring the full data set
when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser might also dynamically create new rules in response to certain
inputs, e.g.\ diagnostic messages indicating that the program which
produced the input being parsed had read a new configuration file, and had
changed its behaviour appropriately.  These avenues of research and
development have not been pursued by the author, but but the architecture
allows them to easily be undertaken independently.

XXX ATTACHING CONDITIONS TO RULES\@; GIVE PLP EXAMPLE AND REFER TO
IMPLEMENTATION AND RESULTS SECTIONS\@?

XXX STATEFUL PARSING ENABLED BY CONDITIONS, E.G\@.  PARSING C COMMENTS\@.

XXX PRIORITY ATTRIBUTE AND RULE ORDERING\@.

When adding new rules, the rule author must be aware that the new rule may
overlap with an existing rule, i.e.\ some inputs could be parsed by more
than one rule.  The architecture does not try to detect overlapping rules:
that responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each input might change between parser invocations.  Overlapping
rules are frequently a requirement, allowing a more specific rule to match
some inputs and a more general rule to match the remainder, e.g.\
separating \acronym{SMTP} delivery to specific sites from \acronym{SMTP}
delivery to the rest of the world.  Allowing overlapping rules simplifies
both the general rule and the more specific rule; additionally rules from
different sources can be combined with a minimum of prior cooperation or
modification required.  Overlapping rules should have a priority attribute
to specify their relative ordering; negative priorities may be useful for
catchall rules.

XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
PARAGRAPH\@.

Decoupling the rules from the actions allows external tools to be written
to detect overlapping rules.  Traditional regexes are equivalent in
computational power to \acronym{FA} and can be converted to \acronym{FA},
so regex overlap can be detected by finding a non-empty intersection of two
\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
for example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
computation complexity XXX MENTION DEMORGAN''S LAW\@?  COMMENT ON THE
COMPUTATION COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE
SPINNING WHEEL BOOK\@.  Perl 5.10 regexes are more powerful than
traditional regexes: it is possible to match correctly balanced brackets
nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are thus
equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
detecting overlap may require calculating the intersection of two
\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
language cannot always be parsed by a \acronym{CFL}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
complexity of calculating \acronym{FA} or \acronym{PDA} intersection.
This is certainly not a task to be performed every time the parser runs:
naive detection of overlap amongst the Postfix log parser's
\numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.  When detecting overlap any
conditions attached to rules must be taken into account, because two rules
whose regexes overlap may have conditions attached which prevent the
overlap in practice.  A less naive approach to overlap detection would
first check for overlapping conditions amongst rules, and then check for
overlapping regexes amongst the rules in each group.  Depending on the
conditions employed a rule may fall into more than one group, e.g.\ given
three conditions:

\begin{enumerate}

    \item $total < 10$

    \item $total > 20$

    \item $total < 30$

\end{enumerate}

\noindent{}The first and second conditions do not overlap, but the third
condition does overlap with both the first and second conditions.  If the
rules are divided into sets based on how their conditions overlap, the
complexity of detecting overlap amongst $n$ rules is:

$$O(n^{2}y)~+~\sum{O(|s|^{2}x)~\forall{}~s~\in{}~overlaps}$$

where:

\begin{tabular}[]{rcl}

            y & = & cost of checking for overlap between two conditions \\
            x & = & cost of checking for overlap between two regexes    \\
     overlaps & = & set of sets of rules with overlapping conditions    \\

\end{tabular}

It is possible to define pathological regexes which fall into two main
categories: regexes that match every input, and regexes that consume
excessive amounts of CPU time during matching.  Defining a regex to match
all inputs is trivial: \verb!/^/! matches the start of every input.  This
regex would be highlighted by a tool that detects overlapping rules, and
would easily be noticed by visual inspection, but more complex regexes
would be harder to spot.  XXX MAYBE GIVE AN EXAMPLE REGEX\@?  Usually
excessive CPU time is consumed when a regex with a lot of alteration and
variable quantifiers fails to match, but successful matching is generally
quite fast (see~\cite{mastering-regular-expressions} for in-depth
discussion).  XXX THIS NEEDS TO BE EXPANDED, OR MOVED TO FIT IN
ELSEWHERE\@.

@@

XXX IMPROVE THIS\@; SHOULD IT BE A SUBSECTION\@?  EXPAND ON ALL THESE, OR
MOVE THEM TO FIT IN ELSEWHERE\@.  ADD MORE IF POSSIBLE\@.

@@

    \item The first matching rule wins: no further rules are tried against
        that input, but there is a mechanism for prioritising the rules so
        that more specific rules can be tried first.  XXX MOVE THIS TO THE
        PRIORITY SECTION\@.

    \item Rules are completely self-contained and can be understood in
        isolation, without reference to any other rules.  This is similar
        to a \acronym{CFG}; in \acronym{CFG} terms the parser rules could
        be described as:

        $\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

@@

XXX EXPAND ON ALL THESE\@.

@@

    \item [Matching rules against inputs is simple:]  The first matching
        rule determines the action that will be invoked: there is no
        backtracking to try alternate rules; no attempt is made to pick a
        \textit{best\/} rule.

    \item [Line oriented:]  The architecture is line oriented at present:
        there is no facility for rules to consume more input or push unused
        input back onto the input stream.  The current input can be
        manipulated by actions, using the cascaded parsing feature.  This
        was not a deliberate design decision, rather a consequence of the
        line oriented nature of Postfix log files; more flexible approaches
        could be pursued.

    \item [Context-free rules:]  Rules can not take into account past or
        future inputs when classifying the current input, though conditions
        can affect which rules will be used to classify each input.

@@

    \item [Transduction:]  The architecture can be thought of as
        implementing transduction: it takes data in one form
        and transforms it to another form; \parsername{} transforms log
        files to a database.

    \item [Similarity to Natural Language Processing:] Unlike traditional
        parsers such as those used when compiling a programming language,
        this architecture does not require a fixed grammar specification
        that inputs must adhere to.  The architecture is capable of dealing
        with interleaved inputs, out of order inputs, and ambiguous inputs
        where heuristics must be applied --- all have arisen and been
        successfully accommodated in the Postfix log file parser.
